# Area of polygons and sampling

```{r, echo=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(
	fig.width = 5,
	fig.height = 5,
	fig.cap = '',
	collapse = TRUE
)
```

## Area and perimeter of polygons

You can compute the area and perimeter of spherical polygons like this. 

```{r}
pol <- rbind(c(-120,-20), c(-80,5), c(0, -20), c(-40,-60), c(-120,-20))
areaPolygon(pol)
perimeter(pol)
```

## Sampling longitude/latitude data


Random or regular sampling of longitude/latitude values on the globe needs to consider that the globe is spherical. That is, if you would take random points for latitude between -90 and 90 and for longitude between -180 and 180, the density of points would be higher near the poles than near the equator.
In contrast, functions 'randomCoordinates' and 'randomCoordinates' return samples that are spatially balanced.

```{r, geo18}
plot(wrld, type='l', col='grey')
a = randomCoordinates(500)
points(a, col='blue', pch=20, cex=0.5)
b = regularCoordinates(3)
points(b, col='red', pch='x')
```


