# 2. The length of a coastline

```{r, echo=FALSE, include=FALSE}
library(rgdal)
```

*How Long Is the Coast of Britain? Statistical Self-Similarity and Fractional Dimension* is the title of a [famous paper](https://classes.soe.ucsc.edu/ams214/Winter09/foundingpapers/Mandelbrot1967.pdf) by [Benoît Mandelbrot](https://en.wikipedia.org/wiki/Benoit_Mandelbrot). Mandelbrot uses data from a paper by [Lewis Fry Richardson](https://en.wikipedia.org/wiki/Lewis_Fry_Richardson) who showed that the length of a coastline changes with scale, or, more precisely, with the length (resolution) of the measuring stick (ruler) used. Mandelbrot discusses the fractal dimension *D* of such lines. *D* is 1 for a straight line, and higher for more wrinkled shapes. For the west coast of Britain, Mandelbrot reports that *D*=1.25. Here I show how to measure the length of a coast line with rulers of different length and how to compute a fractal dimension. 

First we get a high spatial resolution (30 m) coastline for the United Kingdom from the [GADM](http://www.gadm.org) database.

```{r, frac1}
library(raster)
uk <- raster::getData('GADM', country='GBR', level=0)
par(mai=c(0,0,0,0))
plot(uk)
```

This is a single 'multi-polygon' (it has a single feature) and a longitude/latitude coordinate reference system. 
```{r}
data.frame(uk)
```

Let's transform this to a planar coordinate system. That is not required, but it will speed up computations. We used a the Transverse Mercator (tmerc) projection similar to the "British National Grid".  

```{r}
prj <- "+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +datum=WGS84 +units=m"
```
Note that the units are meters.

With that we can transform the coordinates of `uk` from longitude latitude to the British National Grid.

```{r}
library(rgdal)
guk <- spTransform(uk, CRS(prj)) 
```

We only want the main island, so want need to separate (disaggregate) the different polygons.

```{r}
duk <- disaggregate(guk)
head(duk)
```

Now we have 920 features. We want the largest one. 
```{r}
a <- area(duk)
i <- which.max(a)
a[i] / 1000000
b <- duk[i,]
```

Britain has an area of about 220,000 km^2^.

```{r, frac5}
par(mai=rep(0,4))
plot(b)
```

On to the tricky part. The function to go around the coast with a ruler (yardstick) of a certain length.

```{r}
measure_with_ruler <- function(pols, length, lonlat=FALSE) {
	# some sanity checking
	stopifnot(inherits(pols, 'SpatialPolygons'))
	stopifnot(length(pols) == 1)
	
	# get the coordinates of the polygon
	g <- geom(pols)[, c('x', 'y')]
	nr <- nrow(g)
	
	# we start at the first point
	pts <- 1
	newpt <- 1
	while(TRUE) {
	    # start here
		p <- newpt
		
		# order the points 
		j <- p:(p+nr-1)
		j[j > nr] <- j[j > nr] - nr
		gg <- g[j,]
		
		# compute distances
		pd <- pointDistance(gg[1,], gg, lonlat)
		
		# get the first point that is past the end of the ruler
		# this is precise enough for our high resolution coastline
		i <- which(pd > length)[1]
		if (is.na(i)) {
			stop('Ruler is longer than the maximum distance found')
		}
		
		# get the record number for new point in the original order 
		newpt <- i + p
		
		# stop if past the last point
		if (newpt >= nr) break
		
		pts <- c(pts, newpt)
	}
	# add the last (incomplete) stick.
	pts <- c(pts, 1)
	# return the locations
	g[pts, ]
}
```

Now we have the function, life is easy, we just call it a couple of times, using rulers of different lengths. 

```{r}
y <- list()
rulers <- c(25,50,100,150,200,250) # km
for (i in 1:length(rulers)) {
	y[[i]] <- measure_with_ruler(b, rulers[i]*1000)
}
```

Object `y` is a list of matrices containing the locations where the ruler touched the coast. We can plot these on top of a map of Britain.

```{r, frac15, fig.width = 8, fig.height = 10}
par(mfrow=c(2,3), mai=rep(0,4))
for (i in 1:length(y)) {
	plot(b, col='lightgray', lwd=2)
	p <- y[[i]]
	lines(p, col='red', lwd=3)
	points(p, pch=20, col='blue', cex=2)
	
    bar <- rbind(cbind(525000, 900000), cbind(525000, 900000-rulers[i]*1000))
	lines(bar, lwd=2)
	points(bar, pch=20, cex=1.5)
	text(525000, mean(bar[,2]), paste(rulers[i], '  km'), cex=1.5)
	text(525000, bar[2,2]-50000, paste0('(', nrow(p), ')'), cex=1.25)
}
```

*The coastline of Britain, measured with rulers of different lengths. The number of segments is in parenthesis.*
f

Here is the fractal (log-log) plot. Note how the axes are on the log scale, but that I used the non-transformed values for the labels. 

```{r, frac20, fig.width=6, fig.height=6}
# number of times a ruler was used
n <- sapply(y, nrow)

# set up empty plot
plot(log(rulers), log(n), type='n', xlim=c(2,6), ylim=c(2,6), axes=FALSE,
       xaxs="i",yaxs="i", xlab='Ruler length (km)', ylab='Number of segments')

# axes
tics <- c(1,10,25,50,100,200,400)
axis(1, at=log(tics), labels=tics)
axis(2, at=log(tics), labels=tics, las=2)

# linear regression line
m <- lm(log(n)~log(rulers))
abline(m, lwd=3, col='lightblue')

# add observations
points(log(rulers), log(n), pch=20, cex=2, col='red')
```

What does this mean? Let's try some very small rulers, from 1 mm to 10 m.

```{r fracplot, fig.width=6, fig.height=6}
small_rulers <- c(0.000001, 0.00001, 0.0001, 0.001, 0.01)  # km
nprd <- exp(predict(m, data.frame(rulers=small_rulers)))
coast <- nprd * small_rulers
plot(small_rulers, coast, xlab='Length of ruler', ylab='Length of coast', pch=20, cex=2, col='red')
```

So as the ruler get smaller, the coastline gets exponentially longer. As the ruler approaches zero, the length of the coastline approaches infinity. 

The fractal dimension *D* of the coast of Britain is the (absolute value of the) slope of the regression line.

```{r}
m
```

Get the slope 
```{r}
-1 * m$coefficients[2]
```

Very close to Mandelbrot's *D* = 1.25 for the west coast of Britain.

[Further reading](http://www.wahl.org/fe/HTML_version/link/FE4W/c4.htm).
