# 10. Kriging

```{r, include=FALSE}
library(knitr)
opts_chunk$set(
  fig.width  = 5,
  fig.height = 5,
  fig.cap = '',
  collapse = TRUE
)
library(rgdal)
library(raster)
library(gstat)
library(fields)
```


## 10.1 Alberta Rainfall

Recreating Figures 10.2, 10.13 & 10.14 in [O'Sullivan and Unwin (2010)](http://www.wiley.com/WileyCDA/WileyTitle/productCd-0470288574.html). 

### Get the data

You can download the data for this lesson [here](https://biogeo.ucdavis.edu/data/rspatial/alberta.csv) or with the code below.

```{r getData, echo=TRUE}
# Make sure the data directory exists. If not, make it
dir.create('data', showWarnings = FALSE)
files <- c('alberta.csv')
for (filename in files) {
	localfile <- file.path('data', filename)
	if (!file.exists(localfile)) {
		download.file(paste0('https://biogeo.ucdavis.edu/data/rspatial/', filename), dest=localfile)
	}
}
```

### Lesson

Figure 10.2
```{r, krig1}
a <- read.csv('data/alberta.csv')

m <- lm(z ~ x + y, data=a)
summary(m)
plot(a[,2:3], xlim=c(0,60), ylim=c(0,60), las=1, pch=20, yaxs="i", xaxs="i")
text(a[,2:3], labels=a$z, pos=4)

# make the contour lines
x <- seq(0, 60, 1)
y <- seq(0, 60, 1)
# all combinations of x and y
xy <- data.frame(expand.grid(x=x,y=y))
z <- predict(m, xy)
z <- matrix(z, 61, 61)
contour(x, y, z, add=TRUE, labcex=1.25)
```

On to distances. First get a distance matrix for locations
```{r}
library(raster)
dp <- pointDistance(a[,2:3], lonlat=FALSE)
dim(a)
dim(dp)
dp[1:5, 1:5]
diag(dp) <- NA
```

Now the distance matrix for the values observated at the locations. 'dist' makes a symmetrical distance matrix that includes each pair only once. The pointDistance function used above returns the distance between each pair twice. To illustrate this:

```{r}
dist(a$z[1:3])
dz <- dist(a$z)
```

We can transform matrix dp to a distance matrix like this
```{r}
dp <- as.dist(dp)
```

Plot a point cloud of spatial distance against the semivariance (Figure 10.13).
```{r, krig5}
# semivariance
semivar <- dz^2 / 2

plot(dp, semivar, xlim=c(0, 80), ylim=c(0,220), xlab=c('Distance between locations'), 
     ylab=c('Semivariance'), pch=20, axes=FALSE, xaxs="i")
axis(1, at=seq(0,80,10))
axis(2, las=1)
```

And plotting semivariance in bins 
```{r, krig10}
# choose a bin width (in spatial distance)
binwidth <- 8
# assign a lag (bin number) to each record 
lag <- floor(dp/binwidth) + 1
# average value for each lag
lsv <- tapply(semivar, lag, mean)
# compute the average distance for each lag
dlag <- tapply(dp, lag, mean)

plot(dlag, lsv, pch=20, axes=FALSE, xlab='Distance', ylab='Semivariance', xlim=c(0,80))
axis(1, at=seq(0,80,10))
axis(2, las=1)
```

-> continue to analysis / interpolation
