# Local statistics

```{r, echo=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(
  fig.width = 8, fig.height = 5,  fig.cap='',  collapse = TRUE
)
opts_knit$set(
  progress = FALSE,
  global.par = TRUE
)

library(spdep)
library(rspatial)
```

## Introduction

This handout accompanies Chapter 8 in [O'Sullivan and Unwin (2010)](http://www.wiley.com/WileyCDA/WileyTitle/productCd-0470288574.html). 


## LISA

### Get the data

Read the Auckland data.

```{r, loca1}
if (!require("rspatial")) devtools::install_github('rspatial/rspatial')
library(rspatial)
auck <- sp_data('auctb.rds')
``` 

Local Getis *Gi*

```{r, loca2}
library(spdep)
wr <- poly2nb(auck, row.names=auck$Id, queen=FALSE)
lstw <- nb2listw(wr, style='B')
Gi <- localG(auck$TB, lstw)
head(Gi)
```


```{r, loca3}
par(mai=c(0,0,0,0))
Gcuts <- cut(Gi, 5)
Gcutsi <- as.integer(Gcuts)
cols <- rev(gray(seq(0,1,.2)))
plot(auck, col=cols[Gcutsi])
legend('bottomleft', levels(Gcuts), fill=cols)
```

To get the *Gi* * we need to include each polygon as its own 'neighbor'

```{r, loca4}
ws <- include.self(wr)
lstws <- nb2listw(ws, style='B')
Gis <- localG(auck$TB, lstws)
Gscuts <- cut(Gis, 5)
Gscutsi <- as.integer(Gscuts)
cols <- rev(gray(seq(0,1,.2)))
plot(auck, col=cols[Gscutsi])
legend('bottomleft', levels(Gscuts), fill=cols)
```


This looks very similar to the local average.

```{r, loca6}
m <- sapply(ws, function(i) mean(auck$TB[i]))
cts <- cut(m, 5)
mcts <- as.integer(cts)
plot(auck, col=cols[mcts])
legend('bottomleft', levels(cts), fill=cols)
```


The local Moran *Ii* shows where there are locally high levels of autocorrelation.

```{r, loca8}
Ii <- localmoran(auck$TB, lstw)
Icuts <- cut(Ii, 5)
Icutsi <- as.integer(Icuts)
plot(auck, col=cols[Icutsi])
legend('bottomleft', levels(Icuts), fill=cols)
```

## Geographically weighted regression

Here is an example of GWR with California precipitation data (you can download the data with the scripts or links on the top of this page). 


```{r, loca10}
cts <- sp_data('counties.rds')
p <- sp_data('precipitation.csv')
head(p)

plot(cts)
points(p[,c('LONG', 'LAT')], col='red', pch=20)
```

Compute annual average precipitation
```{r, loca11}
p$pan <- rowSums(p[,6:17])
```

Global regression model
```{r, loca12}
m <- lm(pan ~ ALT, data=p)
m
```

Create `Spatial*` objects with a planar crs.

```{r, loca13}
alb <- CRS("+proj=aea +lat_1=34 +lat_2=40.5 +lat_0=0 +lon_0=-120 +x_0=0 +y_0=-4000000 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")
sp <- p
coordinates(sp) = ~ LONG + LAT
crs(sp) <- "+proj=longlat +datum=NAD83"
spt <- spTransform(sp, alb)
ctst <- spTransform(cts, alb)
```


Get the optimal bandwidth
```{r, loca14}
library( spgwr )
bw <- gwr.sel(pan ~ ALT, data=spt)
bw
``` 

Create a regular set of points to estimate parameters for.
```{r, loca16}
r <- raster(ctst, res=10000)
r <- rasterize(ctst, r)
newpts <- rasterToPoints(r)
```


Run the `gwr` function
```{r, loca17}
g <- gwr(pan ~ ALT, data=spt, bandwidth=bw, fit.points=newpts[, 1:2])
g
```

Link the results back to the raster
```{r, loca18, fig.width=9}
slope <- r
intercept <- r
slope[!is.na(slope)] <- g$SDF$ALT 
intercept[!is.na(intercept)] <- g$SDF$'(Intercept)'
s <- stack(intercept, slope)
names(s) <- c('intercept', 'slope')
plot(s)
```

See [this page](http://rspatial.org/analysis/6-local_regression.html) for a more detailed example of Geographically weighted regression.
