# Spatial autocorrelation

```{r, echo=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(fig.width = 5, fig.height = 5, fig.cap='', collapse = TRUE )
library(rgeos)
library(raster)
library(spdep)
library(deldir)
```

## 7.1 Introduction

This handout accompanies Chapter 7 in [O'Sullivan and Unwin (2010)](http://www.wiley.com/WileyCDA/WileyTitle/productCd-0470288574.html). 

### Get the data

```{r getData, echo=TRUE}
if (!require("rspatial")) devtools::install_github('rspatial/rspatial')
library(rspatial)
```


## 7.2 Area of a polygon

Create a polygon like in Figure 7.2 (page 192).

```{r}
pol <- matrix(c(1.7, 2.6, 5.6, 8.1, 7.2, 3.3, 1.7, 4.9, 7, 7.6, 6.1, 2.7, 2.7, 4.9), ncol=2)
library(raster)
sppol <- spPolygons(pol)
```

For illustration purposes, we create the "negative area" polygon as well
```{r}
negpol <- rbind(pol[c(1,6:4),], cbind(pol[4,1], 0), cbind(pol[1,1], 0))
spneg <- spPolygons(negpol)
```

Now plot
```{r, polygons}

cols <- c('light gray', 'light blue')
plot(sppol, xlim=c(1,9), ylim=c(1,10), col=cols[1], axes=FALSE, xlab='', ylab='', 
      lwd=2, yaxs="i", xaxs="i")
plot(spneg, col=cols[2], add=T)
plot(spneg, add=T, density=8, angle=-45, lwd=1)
segments(pol[,1], pol[,2], pol[,1], 0)
text(pol, LETTERS[1:6], pos=3, col='red', font=4)
arrows(1, 1, 9, 1, 0.1, xpd=T)
arrows(1, 1, 1, 9, 0.1, xpd=T)
text(1, 9.5, 'y axis', xpd=T)
text(10, 1, 'x axis', xpd=T)
legend(6, 9.5, c('"positive" area', '"negative" area'), fill=cols, bty = "n")
``` 
 

Compute area
```{r}
p <- rbind(pol, pol[1,])
x <- p[-1,1] - p[-nrow(p),1]
y <- (p[-1,2] + p[-nrow(p),2]) / 2
sum(x * y)
```

Or simply use an existing function

```{r}
# make sure that the coordinates are interpreted as planar (not longitude/latitude)
crs(sppol) <- '+proj=utm +zone=1'
area(sppol)
``` 

## 7.3 Contact numbers

"Contact numbers" for the lower 48 states. Get the polygons:
```{r}
library(raster)
usa <- raster::getData('GADM', country='USA', level=1)
usa <- usa[! usa$NAME_1 %in% c('Alaska', 'Hawaii'), ]
```

To find adjacent polygons, we can use the `spdep` package.

```{r, message=FALSE}
library(spdep)
``` 

We use `poly2nb` to create a neighbors-list. And from that a neighbors matrix.

```{r}
# patience, this takes a while: 
wus <- poly2nb(usa, row.names=usa$OBJECTID, queen=FALSE)
wus
wmus <- nb2mat(wus, style='B', zero.policy = TRUE)
dim(wmus)
```
Compute the number of neighbors for each state.

```{r}
i <- rowSums(wmus)
round(100 * table(i) / length(i), 1)
```

Apparantly, I am using a different data set than OSU (compare the above with table 7.1). By changing the `level` argument to `2` in the `getData` function you can run the same for counties. Which county has 13 neighbors?

## 7.4 Spatial structure


Read the Auckland data.

```{r, auck1}
if (!require("rspatial")) devtools::install_github('rspatial/rspatial')
library(rspatial)
pols <- sp_data("auctb.rds")
``` 


I did not have the tuberculosis data so I guestimated them from figure 7.7. Compare:

```{r, auck2, fig.width=8}
par(mai=c(0,0,0,0))
classes <- seq(0,450,50)
cuts <- cut(pols$TB, classes)
n <- length(classes)
cols <- rev(gray(0:n / n))
plot(pols, col=cols[as.integer(cuts)])
legend('bottomleft', levels(cuts), fill=cols)
```


Create a Rooks' case neighborhood object.
```{r}
wr <- poly2nb(pols, row.names=pols$Id, queen=FALSE)
class(wr)
summary(wr)
``` 

Plot the links between the polygons.
```{r, links, fig.width=6}
par(mai=c(0,0,0,0))
plot(pols, col='gray', border='blue')
xy <- coordinates(pols)
plot(wr, xy, col='red', lwd=2, add=TRUE)
``` 

We can create a matrix from the links list.
```{r}
wm <- nb2mat(wr, style='B')
dim(wm)
```
 
And inspect the content or `wr` and `wm`

```{r}
wr[1:6]
wm[1:6,1:11]
``` 

__Question 1__:*Explain the meaning of the first lines returned by wr[1:6])*


Now let's recreate Figure 7.6 (page 202).

We already have the first one (Rook's case adjacency, plotted above). Queen's case adjacency:
```{r}
wq <- poly2nb(pols, row.names=pols$Id, queen=TRUE)
``` 

Distance based:
```{r}
wd1 <- dnearneigh(xy, 0, 1000)
wd25 <- dnearneigh(xy, 0, 2500)
```

Nearest neighbors:
```{r}
k3 <- knn2nb(knearneigh(xy, k=3, RANN=FALSE))
k6 <- knn2nb(knearneigh(xy, k=6, RANN=FALSE))
```

Delauny:
```{r}
library(deldir)
d <- deldir(xy[,1], xy[,2], suppressMsge=TRUE)
```

Lag-two Rook:
```{r}
wr2 <- wr
for (i in 1:length(wr)) {
	lag1 <- wr[[i]]
	lag2 <- wr[lag1]
	lag2 <- sort(unique(unlist(lag2)))
	lag2 <- lag2[!(lag2 %in% c(wr[[i]], i))] 
	wr2[[i]] <- lag2
}
```

And now we plot them all using the `plotit` function. 

```{r, weights, fig.height=12, fig.width=9}
  plotit <- function(nb, lab='') {
  plot(pols, col='gray', border='white')
  plot(nb, xy, add=TRUE, pch=20)
  text(2659066, 6482808, paste0('(', lab, ')'), cex=1.25)
}

par(mfrow=c(4, 2), mai=c(0,0,0,0))
plotit(wr, 'i')
plotit(wq, 'ii')
plotit(wd1, 'iii')
plotit(wd25, 'iv')
plotit(k3, 'v')
plotit(k6, 'vi')
plot(pols, col='gray', border='white')
plot(d, wlines='triang', add=TRUE, pch=20)
text(2659066, 6482808, '(vii)', cex=1.25)
plotit(wr2, 'viii')
```

## 7.5 Moran's *I*

Below I compute Moran's index according to formula 7.7 on page 205 of OSU.

$$
I = \frac{n}{\sum_{i=1}^n (y_i - \bar{y})^2} \frac{\sum_{i=1}^n \sum_{j=1}^n w_{ij}(y_i - \bar{y})(y_j - \bar{y})}{\sum_{i=1}^n \sum_{j=1}^n w_{ij}}
$$


The number of observations
```{r}
n <- length(pols)
``` 

Values 'y' and 'ybar' (the mean of y).

```{r}
y <- pols$TB
ybar <- mean(y)
``` 

Now we need 

$$ 
(y_i - \bar{y})(y_j - \bar{y}) 
$$ 

That is, (yi-ybar)(yj-ybar) for all pairs. 
I show two methods to compute that.

Method 1:
```{r}
dy <- y - ybar
g <- expand.grid(dy, dy)
yiyj <- g[,1] * g[,2]
``` 

Method 2:
```{r}
yi <- rep(dy, each=n)
yj <- rep(dy)
yiyj <- yi * yj
``` 

Make a matrix of the multiplied pairs
```{r}
pm <- matrix(yiyj, ncol=n)
round(pm[1:6, 1:9])
``` 

And multiply this matrix with the weights to set to zero the value for the pairs that are not adjacent.
```{r}
pmw <- pm * wm
wm[1:6, 1:9]
round(pmw[1:6, 1:9])
``` 

We sum the values, to get this bit of Moran's *I*:

$$
\sum_{i=1}^n \sum_{j=1}^n w_{ij}(y_i - \bar{y})(y_j - \bar{y})
$$

```{r}
spmw <- sum(pmw) 
spmw
``` 

The next step is to divide this value by the sum of weights. That is easy.
```{r}
smw <- sum(wm)
sw  <- spmw / smw
``` 

And compute the inverse variance of y
```{r}
vr <- n / sum(dy^2)
``` 

The final step to compute Moran's *I*
```{r}
MI <- vr * sw
MI
``` 

This is how you can (theoretically) estimate the expected value of Moran's *I*. That is, the value you would get in the absence of spatial autocorrelation. Note that it is not zero for small values of *n*.
```{r}
EI <- -1/(n-1)
EI
``` 


After doing this 'by hand', now let's use the spdep package to compute Moran's *I* and do a significance test. To do this we need to create a `listw` type spatial weights object. To get the same value as above we use `style='B'` to use binary (TRUE/FALSE) distance weights. 

```{r}
ww <- nb2listw(wr, style='B')
ww
``` 


On to the `moran` function. Have a look at ?moran. The function is defined as `moran(y, ww, n, Szero(ww))`. Note the odd arguments `n` and `S0`. I think they are odd, because `ww` has that information. Anyway, we supply them and it works. There probably are cases where it makes sense to use other values.


```{r}
moran(pols$TB, ww, n=length(ww$neighbours), S0=Szero(ww))
```

Note that the global sum ow weights
```{r}
Szero(ww)
``` 
Should be the same as
```{r}
sum(pmw != 0)
``` 

Now we can test for significance. First analytically, using linear regression based logic and assumptions.

```{r}
moran.test(pols$TB, ww, randomisation=FALSE)
``` 

And now using Monte Carlo simulation --- which is the preferred method. In fact, the only good method to use.

```{r}
moran.mc(pols$TB, ww, nsim=99)
``` 


__Question 2__: *How do you interpret these results (the significance tests)?*



__Question 3__: *What would a good value be for `nsim`?*



To make a Moran scatter plot we first get the neighbouring values for each value.
```{r}
n <- length(pols)
ms <- cbind(id=rep(1:n, each=n), y=rep(y, each=n), value=as.vector(wm * y))
``` 

Remove the zeros

```{r}
ms <- ms[ms[,3] > 0, ]
``` 

And compute the average neighbour value
```{r}
ams <- aggregate(ms[,2:3], list(ms[,1]), FUN=mean)
ams <- ams[,-1]
colnames(ams) <- c('y', 'spatially lagged y')
head(ams)
``` 

Finally, the plot.
```{r, ngb}
plot(ams)
reg <- lm(ams[,2] ~ ams[,1])
abline(reg, lwd=2)
abline(h=mean(ams[,2]), lt=2)
abline(v=ybar, lt=2)
``` 

Note that the slope of the regression line:

```{r, ngb2}
coefficients(reg)[2]
```
is almost the same as Moran's *I*.
 

Here is a more direct approach to accomplish the same thing (but hopefully the above makes it clearer how this is actually computed). Note the row standardisation of the weights matrix:

```{r}
rwm <- mat2listw(wm, style='W')
# Checking if rows add up to 1
mat <- listw2mat(rwm)
apply(mat, 1, sum)[1:15]
``` 

Now we can plot

```{r, mplot}
moran.plot(y, rwm)
```  

__Question 4__: *Show how to use the 'geary' function to compute Geary's C*


__Question 5__: *Write your own Monte Carlo simulation test to compute p-values for Moran's *I*, replicating the results we obtained with the function from spdep*. Show a figure similar to Figure 7.9 in OSU.


__Question 6__: *Write your own Geary C function, by completing the function below*

```
gearyC <- ((n-1)/sum(( "----")\^2)) * sum(wm * (" --- ")\^2) / (2 * sum(wm))
```

