# Unsupervised Classification

```{r setup, echo=TRUE, include=FALSE}
library(knitr)
library(terra)
```

In this chapter we explore unsupervised classification. Various unsupervised classification algorithms exist, and the choice of algorithm can affect the results. We will explore only one algorithm (k-means) to illustrate the general principle.

For this example, we will follow the [National Land Cover Database 2011 (NLCD 2011)](https://www.mrlc.gov/nlcd2011.php) classification scheme for a subset of the Central Valley regions. We use cloud-free composite image from [Landsat 5](https://landsat.gsfc.nasa.gov/landsat-5/) with 6 bands.


```{r}
library(terra)
landsat5 <- rast('data/rs/centralvalley-2011LT5.tif')
names(landsat5) <- c('blue', 'green', 'red', 'NIR', 'SWIR1', 'SWIR2')
```

__Question 1__: *Make a 3-band False Color Composite plot of `landsat5`.*


In unsupervised classification, we use the reflectance data, but we don't supply any response data (that is, we do not identify any pixel as belonging to a particular class). This may seem odd, but it can be useful when we don't have much prior knowledge of a study area. Or if you have broad knowledge of the distribution of land cover classes of interest, but no specific ground data. 

The algorithm groups pixels with similar spectral characteristics into groups. 

Learn more about K-means and other unsupervised-supervised algorithms [here](http://nptel.ac.in/courses/105108077/module5/lecture19.pdf).

We will perform unsupervised classification on a spatial subset of the `ndvi` layer. Here is yet another way to compute ndvi. In this case we do not use a separate function, but we use a direct algebraic notation.

```{r}
ndvi <- (landsat5[['NIR']] - landsat5[['red']]) / (landsat5[['NIR']] + landsat5[['red']])
```

We will do `kmeans` clustering of the `ndvi` data. First we use `crop` to make a spatial subset of the `ndvi`, to allow for faster processing (you can select any `extent` using the `drawExtent()` function).  

## kmeans classification
```{r kmeans, fig.width = 8, fig.height = 4}
# Extent to crop ndvi layer
e <- ext(-121.807, -121.725, 38.004, 38.072)

# crop landsat by the extent
ndvi <- crop(ndvi, e)
ndvi

# convert the raster to vecor/matrix
nr <- values(ndvi)
str(nr)
```

Please note that `getValues` converted the `ndvi` RasterLayer to an array (matrix). Now we will perform the `kmeans` clustering on the matrix and inspect the output. 

```{r kmeansobject}
# It is important to set the seed generator because `kmeans` initiates the centers in random locations
set.seed(99)

# We want to create 10 clusters, allow 500 iterations, start with 5 random sets using "Lloyd" method
kmncluster <- kmeans(na.omit(nr), centers = 10, iter.max = 500, nstart = 5, algorithm="Lloyd")

# kmeans returns an object of class "kmeans"
str(kmncluster)
```

`kmeans` returns an object with 9 elements. The length of the `cluster` element within `kmncluster` is `r length(kmncluster$cluster)` which same as length of `nr` created from the `ndvi`. The cell values of `kmncluster$cluster` range between 1 to 10 corresponding to the input number of cluster we provided in the `kmeans` function. `kmncluster$cluster` indicates the cluster label for corresponding pixel. We need to convert the `kmncluster$cluster` values back to RasterLayer of the same dimension as the `ndvi`.

```{r kmeansraster}
# Use the ndvi object to set the cluster values to a new raster
knr <- ndvi
values(knr) <- kmncluster$cluster
knr
```

We can see that `knr` is a RasterLayer but we do not know which cluster (1-10) belongs to what land cover class (and if it does belong to a class that we would recognize). You can find that out by plotting them side-by-side with a reference layers and using unique color for each cluster.

```{r kmeansplot, fig.height=4, fig.width=8}

# Define a color vector for 10 clusters (learn more about setting the color later)
mycolor <- c("#fef65b","#ff0000", "#daa520","#0000ff","#0000ff","#00ff00","#cbbeb5",
             "#c3ff5b", "#ff7373", "#00ff00", "#808080")

par(mfrow = c(1,2))
plot(ndvi, col = rev(terrain.colors(10)), main = 'Landsat-NDVI')
plot(knr, main = 'Unsupervised classification', col = mycolor )
```

While for other purposes it is usually better to define more classes (and possibly merge classes later), a simple classification like this one could be useful, e.g., merge cluster 4 and 5 to construct a water mask for the year 2011.

You can change the colors in my `mycolor`. Learn more about selecting colors in R [here](http://www.melissaclarkson.com/resources/R_guides/documents/colors_Ver2.pdf) and [here](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf).

__Question 2__:*Plot 3-band RGB of `landsat5` for the subset (extent `e`) and result of `kmeans` clustering side-by-side and make a table of land-use land-cover labels for the clusters. E.g. cluster 4 and 5 are water.* 

