# High-level methods

Several 'high level' methods (functions) have been implemented for `RasterLayer` objects. 'High level' refers to methods that you would normally find in a GIS program that supports raster data. Here we briefly discuss some of these. See the help files for more detailed descriptions.

The high-level methods have some arguments in common. The first argument is typically 'x' or 'object' and can be a `RasterLayer`, or, in most cases, a `RasterStack` or `RasterBrick`. It is followed by one or more arguments specific to the method (either additional `RasterLayer` objects or other arguments), followed by a filename="" and "..." arguments. 

The default filename is an empty character "". If you do not specify a filename, the default action for the method is to return a `raster` object that only exists in memory. However, if the method deems that the `raster` object to be created would be too large to hold memory it is written to a temporary file instead.

The "..."  argument allows for setting additional arguments that are relevant when writing values to a file: the file format, datatype (e.g. integer or real values), and a to indicate whether existing files should be overwritten. 


## Modifying a Raster* object

There are several methods that deal with modifying the spatial extent of `Raster*` objects. The `crop` method lets you take a geographic subset of a larger `raster` object. You can crop a `Raster*` by providing an extent object or another spatial object from which an extent can be extracted (objects from classes deriving from Raster and from Spatial in the sp package). An easy way to get an extent object is to plot a `RasterLayer` and then use ` drawExtent` to visually determine the new extent (bounding box) to provide to the crop method. 

`trim` crops a `RasterLayer` by removing the outer rows and columns that only contain `NA` values. In contrast, `extend` adds new rows and/or columns with `NA` values. The purpose of this could be to create a new `RasterLayer` with the same Extent of another larger `RasterLayer` such that the can be used together in other methods. 

The `merge` method lets you merge 2 or more ```Raster*``` objects into a single new object. The input objects must have the same resolution and origin (such that their cells neatly fit into a single larger raster). If this is not the case you can first adjust one of the ```Raster*``` objects with use `(dis)aggregate` or `resample`.

`aggregate` and `disaggregate} allow for changing the resolution (cell size) of a ```Raster*``` object. In the case of `aggregate`, you need to specify a function determining what to do with the grouped cell values (e.g. `mean`). It is possible to specify different (dis)aggregation factors in the x and y direction. `aggregate` and `disaggregate` are the best methods when adjusting cells size only, with an integer step (e.g. each side 2 times smaller or larger), but in some cases that is not possible. 

For example, you may need nearly the same cell size, while shifting the cell centers. In those cases, the `resample` method can be used. It can do either nearest neighbor assignments (for categorical data) or bilinear interpolation (for numerical data). Simple linear shifts of a Raster object can be accomplished with the `shift` method or with the `extent` method. `resample` should not be used to create a Raster* object with much larger resolution. If such adjustments need to be made then you can first use aggregate. 

With the `projectRaster` method you can transform values of ```Raster*``` object to a new object with a different coordinate reference system.

Here are some simple examples:
```{r  raster-5, fig=TRUE , echo=TRUE}
library(raster)
r <- raster()
r[] <- 1:ncell(r)
ra <- aggregate(r, 10)
r1 <- crop(r, extent(-180,0,0,30))
r2 <- crop(r, extent(-10,180,-20,10))
m <- merge(r1, r2, filename='test.grd', overwrite=TRUE)
plot(m)
```

`bf` lets you flip the data (reverse order) in horizontal or vertical direction -- typically to correct for a 'communication problem' between different R packages or a misinterpreted file. `rotate` lets you rotate longitude/latitude rasters that have longitudes from 0 to 360 degrees (often used by climatologists) to the standard -180 to 180 degrees system. With `t` you can rotate a ```Raster*``` object 90 degrees.


## Overlay

The `overlay` method can be used as an alternative to the raster algebra discussed above. Overlay, like the methods discussed in the following subsections provide either easy to use short-hand, or more efficient computation for large (file based) objects. 

With `overlay` you can combine multiple Raster objects (e.g. multiply them). The related method `mask` removes all values from one layer that are `NA` in another layer, and `cover` combines two layers by taking the values of the first layer except where these are `NA`.


## Calc

The `calc` method allows you to do a computation for a single `raster` object by providing a function. If you supply a `RasterLayer`, another `RasterLayer` is returned. If you provide a multi-layer object you get a (single layer) `RasterLayer` if you use a summary type function (e.g. `sum`) but a `RasterBrick` if multiple layers are returned. `stackApply` computes summary type layers for subsets of a `RasterStack` or `RasterBrick`.


## Reclassify
You can use `cut` or `reclassify` to replace ranges of values with single values, or `subs` to substitute (replace) single values with other values. 

```{r  raster-6}
r <- raster(ncol=3, nrow=2)
values(r) <- 1:ncell(r)
getValues(r)
s <- calc(r, fun=function(x){ x[x < 4] <- NA; return(x)} )
as.matrix(s)
t <- overlay(r, s, fun=function(x, y){ x / (2 * sqrt(y)) + 5 } )
as.matrix(t)
u <- mask(r, t)
as.matrix(u)
v <- u==s
as.matrix(v)
w <- cover(t, r)
as.matrix(w)
x <- reclassify(w, c(0,2,1,  2,5,2, 4,10,3))
as.matrix(x)
y <- subs(x, data.frame(id=c(2,3), v=c(40,50)))
as.matrix(y)
```


## Focal

The `focal` method currently only works for (single layer) RasterLayer objects. It uses values in a neighborhood of cells around a focal cell, and computes a value that is stored in the focal cell of the output RasterLayer. The neighborhood is a user-defined a matrix of weights and could approximate any shape by giving some cells zero weight. It is possible to only compute new values for cells that are `NA` in the input RasterLayer.


## Distance

There are a number of distance related methods. `distance` computes the shortest distance to cells that are not `NA`. `pointDistance` computes the shortest distance to any point in a set of points. `gridDistance` computes the distance when following grid cells that can be traversed (e.g. excluding water bodies). `direction` computes the direction towards (or from) the nearest cell that is not `NA`. `adjacency` determines which cells are adjacent to other cells, and `pointDistance` computes distance between points. See the `gdistance` package for more advanced distance calculations (cost distance, resistance distance)

## Spatial configuration

The `clump` method identifies groups of cells that are connected. `boundaries` identifies edges, that is, transitions between cell values. `area` computes the size of each grid cell (for unprojected rasters), this may be useful to, e.g. compute the area covered by a certain class on a longitude/latitude raster. 

```{r  raster-7}
r <- raster(nrow=45, ncol=90)
r[] <- round(runif(ncell(r))*3)
a <- area(r)
zonal(a, r, 'sum')
```


## Predictions
The package has two methods to make model predictions to (potentially very large) rasters. `predict` takes a multilayer raster and a fitted model as arguments. Fitted models can be of various classes, including glm, gam, randomforest, and brt. method `interpolate` is similar but is for models that use coordinates as predictor variables, for example in kriging and spline interpolation.


## Vector to raster conversion

The raster packages supports point, line, and polygon to raster conversion with the `rasterize` method. For vector type data (points, lines, polygons), objects of Spatial* classes defined in the `sp` package are used; but points can also be represented by a two-column matrix (x and y). 

Point to raster conversion is often done with the purpose to analyze the point data. For example to count the number of distinct species (represented by point observations) that occur in each raster cell. `rasterize` takes a `Raster*` object to set the spatial extent and resolution, and a function to determine how to summarize the points (or an attribute of each point) by cell. 

Polygon to raster conversion is typically done to create a `RasterLayer` that can act as a mask, i.e. to set to `NA` a set of cells of a `raster` object, or to summarize values on a raster by zone. For example a country polygon is transferred to a raster that is then used to set all the cells outside that country to `NA`; whereas polygons representing administrative regions such as states can be transferred to a raster to summarize raster values by region. 

It is also possible to convert the values of a `RasterLayer` to points or polygons, using `rasterToPoints` and `rasterToPolygons`. Both methods only return values for cells that are not `NA`. Unlike `rasterToPolygons`, `rasterToPoints` is reasonably efficient and allows you to provide a function to subset the output before it is produced (which can be necessary for very large rasters as the point object is created in memory). 


## Summarize

When used with a ```Raster*``` object as first argument, normal summary statistics functions such as min, max and mean return a RasterLayer. You can use cellStats if, instead, you want to obtain a summary for all cells of a single ```Raster*``` object. You can use `freq` to make a frequency table, or to count the number of cells with a specified value. Use `zonal` to summarize a ```Raster*``` object using zones (areas with the same integer number) defined in a `RasterLayer` and `crosstab` to cross-tabulate two `RasterLayer` objects.

```{r  raster-10}
r <- raster(ncol=36, nrow=18)
r[] <- runif(ncell(r))
cellStats(r, mean)
s = r
s[] <- round(runif(ncell(r)) * 5)
zonal(r, s, 'mean')
freq(s)
freq(s, value=3)
crosstab(r*3, s)
```
