# Unsupervised Image Classification

```{r setup, echo=TRUE, include=FALSE}
library(knitr)
```

In this chapter we explore unsupervised classification. Various unsupervised classification algorithms exist, and the choice of algorithm can affect the results. We will explore only one algorithm (k-means) to illustrate the general principle.

For this example, we will follow the [National Land Cover Database 2011 (NLCD 2011)](https://www.mrlc.gov/nlcd2011.php) classification scheme for a subset of the Central Valley regions. You will use cloud-free composite image from [Landsat 5](https://landsat.gsfc.nasa.gov/landsat-5/) with 6 bands.


```{r}
library(raster)
landsat5 <- stack('data/rs/centralvalley-2011LT5.tif')
names(landsat5) <- c('blue', 'green', 'red', 'NIR', 'SWIR1', 'SWIR2')
```

**Exercise 1** Make a 3-band False Color Composite plot of `landsat5`.


In unsupervised classification, we don't supply any training data. This particularly is useful when we don't have prior knowledge of the study area. The algorithm groups pixels with similar spectral characteristics into unique clusters/classes/groups following some statistically determined conditions (e.g. minimizing mean square root error within each cluster). You have to re-label and combine these spectral clusters into information classes (for e.g. land-use land-cover). Unsupervised algorithms are often referred as clustering.

To get satisfactory results from unsupervised classification, a good practice is to start with large number of centers (more clusters) and merge/group/recode similar clusters by inspecting the original imagery.

Learn more about K-means and other unsupervised-supervised algorithms [here](http://nptel.ac.in/courses/105108077/module5/lecture19.pdf).

We will perform unsupervised classification on a spatial subset of the `ndvi` layer.

```{r}
ndvi <- (landsat5[['NIR']]-landsat5[['red']])/(landsat5[['NIR']]+landsat5[['red']])
```

Now we will perform unsupervised `kmeans` clustering on the `ndvi` layer. First we use `crop` to make a spatial subset of the `ndvi` layer. The smaller spatial extent allows for faster processing (you can select any `extent` using the `drawExtent()` function).  

## kmeans classification
```{r kmeans, fig.width = 8, fig.height = 4}
# Extent to crop ndvi layer
e <- extent(-121.807, -121.725, 38.004, 38.072)

# crop landsat by the extent
ndvi <- crop(ndvi, e)
ndvi

# convert the raster to vecor/matrix
nr <- getValues(ndvi)
str(nr)
```

Please note that `getValues` converted the `ndvi` RasterLayer to array. Now we will perform the `kmeans` clustering on the array object and inspect the output. 

```{r kmeansobject}
# It is important to set the seed generator because `kmeans` initiates the centers in random locations
set.seed(99)

# We want to create 10 clusters, allow 500 iterations, start with 5 random sets using "Lloyd" method
kmncluster <- kmeans(na.omit(nr), centers = 10, iter.max = 500, nstart = 5, algorithm="Lloyd")

# kmeans returns an object of class "kmeans"
str(kmncluster)
```
`kmeans` returns an object with 9 elements. The length of the `cluster` element within `kmncluster` is `r length(kmncluster$cluster)` which same as length of `nr` created from the `ndvi`. The cell values of `kmncluster$cluster` range between 1 to 10 corresponding to the input number of cluster we provided in the `kmeans` function. `kmncluster$cluster` indicates the cluster label for corresponding pixel. We need to convert the `kmncluster$cluster` array back to RasterLayer of the same dimension as the `ndvi`.

```{r kmeansraster}
# First create a copy of the ndvi layer 
knr <- ndvi

# Now replace raster cell values with kmncluster$cluster array
knr[] <- kmncluster$cluster

# Alternative
values(knr) <- kmncluster$cluster
knr
```

We can see that `knr` is a RasterLayer but we do not know which cluster (1-10) belongs what LULC class. You can find that out by plotting them side-by-side with other reference layer and using unique color for each cluster.

```{r kmeansplot, fig.height=4, fig.width=8}

# Define a color vector for 10 clusters (learn more about setting the color later)
mycolor <- c("#fef65b","#ff0000", "#daa520","#0000ff","#0000ff","#00ff00","#cbbeb5",
             "#c3ff5b", "#ff7373", "#00ff00", "#808080")

par(mfrow = c(1,2))
plot(ndvi, col = rev(terrain.colors(10)), main = 'Landsat-NDVI')
plot(knr, main = 'Unsupervised classification', col = mycolor )
```

While for other purposes it is usually better to define more classes (and possibly fuse classes later), a simple classification like this one could be useful, e.g., merge cluster 4 and 5 to construct a water mask for the year 2011.

**Important** These 10 clusters are not useful unless you re-label them with LULC information class. You can save the raster output of clustering and open it in GIS platform (e.g. [QGIS](https://qgis.org/en/site/about/index.html)) and assign classes to each clusters. Multiple clusters can have the same class! A tutorial can be found [here](http://wiki.awf.forst.uni-goettingen.de/wiki/index.php/Unsupervised_classification_(Tutorial)). You can skip step 1 in the tutorial which provides guidelines for generating clusters.

You can change the colors in my `mycolor`. Learn more about selecting colors in R [here](http://www.melissaclarkson.com/resources/R_guides/documents/colors_Ver2.pdf) and [here](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf).

**Exercise** Plot 3-band RGB of `landsat5` for the subset (extent `e`) and result of `kmeans` clustering side-by-side and make a table of land-use land-cover labels for the clusters. E.g. cluster 4 and 5 are water. 

