# Vector data manipulation


```{r, echo=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(
	fig.width = 4,
	fig.height = 5,
	fig.cap = '',
	collapse = TRUE
)

opts_knit$set(
    progress = FALSE,
	global.par = TRUE
)
	
library(raster)
library(rgeos)
library(rgdal)

```

Example SpatialPolygons
```{r, vec1}
f <- system.file("external/lux.shp", package="raster")
library(raster)
p <- shapefile(f)
p
par(mai=c(0,0,0,0))
plot(p)
```

## 7.1 Basics

Basic operations are pretty much like working with a data.frame. 

### 7.1.1 Geometry and attributes

To extracting the attributes (data.frame) from a Spatial object, use:
```{r}
d <- data.frame(p)
head(d)
```

Extracting geometry (rarely needed).
```{r}
g <- geom(p)
head(g)
```

### 7.1.2 Variables

Extracting a variable.
```{r}
p$NAME_2
```

Sub-setting by variable. Note how this is different from the above example. Above a vector of values is returned. With the approach below you get a new SpatialPolygonsDataFrame with only one variable. 
```{r}
p[, 'NAME_2']
```

Adding a new variable.
```{r}
set.seed(0)
p$new <- sample(letters, length(p))
p
```

Assigning a new value to an existing variable.
```{r}
p$new <- sample(LETTERS, length(p))
p
```

To get rid of a variable.

```{r}
p$new <- NULL
```

### 7.1.3 Merge

You can join a table (data.frame) with a Spatial* object with `merge`.

```{r}
dfr <- data.frame(District=p$NAME_1, Canton=p$NAME_2, Value=round(runif(length(p), 100, 1000)))
dfr <- dfr[order(dfr$Canton), ]
pm <- merge(p, dfr, by.x=c('NAME_1', 'NAME_2'), by.y=c('District', 'Canton'))
pm
```


### 7.1.4 Records

Selecting rows (records).
```{r}
i <- which(p$NAME_1 == 'Grevenmacher')
g <- p[i,]
g
```

It is also possible to interactively selecion and query records by clikcing on a plotted dataset. That is difficult to show here. See `?select` for interactively selecting spatial features and `?click` to identify attributes by clicking on a plot (map).


## 7.2 Append and aggregate


## 7.2.1 Append

More example data. Object `z`, consisting of four, and `z2` which is one of these four polygons.

```{r, zzz}
z <- raster(p, nrow=2, ncol=2, vals=1:4)
names(z) <- 'Zone'
# coerce RasterLayer to SpatialPolygonsDataFrame
z <- as(z, 'SpatialPolygonsDataFrame')
z
z2 <- z[2,]
plot(p)
plot(z, add=TRUE, border='blue', lwd=5)
plot(z2, add=TRUE, border='red', lwd=2, density=3, col='red')
```


To append Spatial* objects of the same (vector) type you can use `bind`

```{r}
b <- bind(p, z)
head(b)
tail(b)
```

Note how `bind` allows you to append `Spatial*` objects with different attribute names. 


## 7.2.2 Aggregate
```{r, agg}
pa <- aggregate(p, by='NAME_1')
za <- aggregate(z)
plot(za, col='light gray', border='light gray', lwd=5)
plot(pa, add=TRUE, col=rainbow(3), lwd=3, border='white')
```

You can also aggregate by providing a second Spatial object (see `?sp::aggregate`)

Aggregate without dissolve
```{r, aggnodis}
zag <- aggregate(z, dissolve=FALSE)
zag
plot(zag, col="light gray")
```

This is a structure that is similar to what you may get for an archipelago: multiple polygons represented as one entity (one row).
Use `disaggregate` to split these up into their parts.

```{r}
zd <- disaggregate(zag)
zd
```


## 7.3 Overlay

### 7.3.1 Erase

Erase a part of a SpatialPolygons object
```{r}
e <- erase(p, z2)
```

This is equivalent to 
```{r, erase}
e <- p - z2
plot(e)
```

### 7.3.2 Intersect

Intersect SpatialPolygons

```{r, int}
i <- intersect(p, z2)
plot(i)
```

This is equivalent to 
```{r}
i <- p * z2
```

You can also intersect with an Extent (rectangle). 

```{r, intext}
e <- extent(6, 6.4, 49.7, 50)
pe <- crop(p, e)
plot(p)
plot(pe, col='light blue', add=TRUE)
plot(e, add=TRUE, lwd=3, col='red')

```


### 7.3.3 Union

Get the union of two SpatialPolygon* objects.

```{r}
u <- union(p, z)
```

This is equivalent to

```{r}
u <- p + z
```

Note that there are many more polygons now. One for each unique combination of polygons (and attributes in this case).
```{r}
u
set.seed(5)
plot(u, col=sample(rainbow(length(u))))
```

### 7.3.4 Cover

Cover is a combination of intersect and union

```{r, cov}
cov <- cover(p, z)
cov
plot(cov)
```


### 7.3.4 Difference


The symmetrical difference of two SpatialPolygons* objects

```{r, dif}
dif <- symdif(z,p)
plot(dif, col=rainbow(length(dif)))
dif
```

## 7.4 Spatial queries

Query polygons with points. 

```{r, pts}
pts <- matrix(c(6, 6.1, 5.9, 5.7, 6.4, 50, 49.9, 49.8, 49.7, 49.5), ncol=2)
spts <- SpatialPoints(pts, proj4string=crs(p))
plot(z, col='light blue', lwd=2)
points(spts, col='light gray', pch=20, cex=6)
text(spts, 1:nrow(pts), col='red', font=2, cex=1.5)
lines(p, col='blue', lwd=2)
```

Use `over` for  queries between Spatial* objects

```{r}
over(spts, p)
over(spts, z)
```

`extract` is generally used for queries between Spatial* and Raster* objects, but it can also be used here.
```{r}
extract(z, pts)
```


