# Raster Advanced 


```{r, echo=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(
	fig.width = 5,
	fig.height = 5,
	fig.cap = '',
	collapse = TRUE
)
```


## Session options

There is a number of session options that influence reading and writing files. These can be set in a session, with `rasterOptions`, and saved to make them persistent in between sessions. But you probably should not change the default values unless you have pressing need to do so. You can, for example, set the directory where temporary files are written, and set your preferred default file format and data type. Some of these settings can be overwritten by arguments to functions where they apply (with arguments like filename, datatype, format). Except for generic functions like mean, '+', and sqrt. These functions may write a file when the result is too large to hold in memory and then these options can only be set through the session options.  The options chunksize and maxmemory determine the maximum size (in number of cells) of a single chunk of values that is read/written in chunk-by-chunk processing of very large files.

```{r, raster-119}
rasterOptions()
```


## Extending raster objects

It is straightforward to build on the Raster* objects using the S4 inheritance mechanism. Say you need objects that behave like a `RasterLayer`, but have some additional properties that you need to use in your own functions (S4 methods). See Chambers (2009) and the help pages of the Methods package for more info. Below is an example:

```{r, raster-132}
setClass ('myRaster',
	contains = 'RasterLayer',
	representation (
		important = 'data.frame',
		essential = 'character'
	) ,
	prototype (
		important = data.frame(),
		essential = ''
	)
)
	
r <- raster(nrow=10, ncol=10)

m <- as(r, 'myRaster')
m@important <- data.frame(id=1:10, value=runif(10))
m@essential <- 'my own slot'
m[] <- 1:ncell(m)
```

```{r, raster-133}
setMethod ('show' , 'myRaster', 
	function(object) {
		callNextMethod(object) # call show(RasterLayer)
		cat('essential:', object@essential, '\n')
		cat('important information:\n')
		print( object@important)
	})	
	
m
```





## Other classes

Below is some more detail, you do not need to read or understand this section to use the `raster` package. 

The three classes described above inherit from the `raster` class (that means they are derived from this more basic 'parent' class by adding something to that class) which itself inherits from the `BasicRaster` class. The `BasicRaster` only has a few properties (referred to as 'slots' in S4 speak): the number of columns and rows, the coordinate reference system (which itself is an object of class `CRS`, which is defined in package '`sp`') and the spatial extent, which is an object of class `Extent`.  

An object of class `Extent` has four slots: xmin, xmax, ymin, and ymax. These represent the minimum and maximum x and y coordinates of the of the Raster object. These would be, for example,  -180, 180, -90, and 90, for a global raster with longitude/latitude coordinates.  Note that raster uses the coordinates of the extremes (corners) of the entire raster (unlike some files/programs that use the coordinates of the center of extreme cells). 

`raster`  is a virtual class. This means that it cannot be instantiated (you cannot create objects from this class). It was created to allow the definition of methods for that class. These methods will be dispatched when called with a descendent of the class (i.e. when the method is called with a `RasterLayer`, `RasterBrick`  or `RasterStack` object as argument). This allows for efficient code writing because many methods are the same for any of these three classes, and hence a single method for `raster` suffices. 

`RasterStackBrick` is a class union of the `RasterStack` and `RasterBrick` class. This is a also a virtual class. It allows defining methods (functions) that apply to both `RasterStack` and `RasterBrick` objects.




