# Vector data

```{r, echo=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(fig.width = 5, fig.height = 5, fig.cap = '', collapse = TRUE)
```

## Introduction

The `terra` package defines a set of *classes* to represent spatial data.  A class defines a particular data type. The `data.frame` is an example of a class. Any particular `data.frame` you create is an *object* (instantiation) of that class. 

The main reason for defining classes is to create a standard representation of a particular data type to make it easier to write functions (known as "methods") for them. See Hadley Wickham's [Advanced R](http://adv-r.had.co.nz/) or John Chambers' [Software for data analysis](http://www.springer.com/us/book/9780387759357) for a detailed discussion of the use of classes in *R*). 

Package `terra` introduces a number of classes with names that start with `Spat`. For vector data, the relevant class is `SpatVector`. These classes represent geometries as well as attributes (variables) describing the geometries. 

It is possible to create `SpatVector` objects from scratch with *R* code. It can be very useful to create small self contained example to illustrate something, for example to ask a question about how to do a particular operation without needing to give access to the real data you are using (which is always cumbersome). But in real life you will read these from a file or database [see Chapter 5](./5-files.html). 

To get started, let's make some SpatVector objects from scratch anyway, using the same data as were used in the previous chapter. 

## Points

```{r}
longitude <- c(-116.7, -120.4, -116.7, -113.5, -115.5, -120.8, -119.5, -113.7, -113.7, -110.7)
latitude <- c(45.3, 42.6, 38.9, 42.1, 35.7, 38.9, 36.2, 39, 41.6, 36.9)
lonlat <- cbind(longitude, latitude)
```

Now create a `SpatVector` object. First load the `terra` package from the library. If this command fails with `Error in library(terra) : there is no package called ‘terra’`, then you need to *install* the package first, with `install.packages("terra")`


```{r}
library(terra)
pts <- vect(lonlat)
```

Let's check what kind of object `pts` is. 
```{r}
class (pts)
```

And what is inside of it 
```{r}
pts
geom(pts)
```

So we see that the object has the coordinates we supplied, but also an `extent`. This spatial extent was computed from the coordinates. There is also a coordinate reference system ("crs", discussed in more detail later). We did not provide the crs so it is unknown (`NA`). That is not good, so let's recreate the object, and now provide a crs. 
 
```{r}
crdref <- "+proj=longlat +datum=WGS84"
pts <- vect(lonlat, crs=crdref)
pts
crs(pts)
```

We can add attributes (variables) to the `SpatVector` object. First we need a `data.frame` with the same number of rows as there are geometries.

```{r}
# Generate random precipitation values, same quantity as points
precipvalue <- runif(nrow(lonlat), min=0, max=100)
df <- data.frame(ID=1:nrow(lonlat), precip=precipvalue)
```

Combine the `SpatVector` with the `data.frame`.


```{r}
ptsdf <- vect(lonlat, data=df)
crs(ptsdf) <- crdref
ptsdf
```

To see what is inside:
```{r}
show(ptsdf)
```

## Lines and polygons

Making a `SpatVector` of points was easy. Making a `SpatVector` of lines or polygons is a bit more complex, but stil relatively straightforward. 

```{r}
lon <- c(-116.8, -114.2, -112.9, -111.9, -114.2, -115.4, -117.7)
lat <- c(41.3, 42.9, 42.4, 39.8, 37.6, 38.3, 37.6) 
lonlat <- cbind(id=1, part=1, lon, lat)
lns <- vect(lonlat, type="lines", crs=crdref)
lns
```


```{r}
pols <- vect(lonlat, type="polygons", crs=crdref)
pols
```

Behind the scenes the class deals with the complexity of accommodate the possibility of multiple polygons, each consisting of multiple sub-polygons, some of which may be "holes".  Fortunately, you do not need to understand how these structures are organized. The main take home message is that a `SpatVector` stores geometries (coordinates), the name of the coordinate reference system, and attributes. 

We can make use `plot` to make a map.

```{r, vectordata-1}
plot(pols, las=1)
plot(pols, border='blue', col='yellow', lwd=3, add=TRUE)
points(pts, col='red', pch=20, cex=3)
```

We'll make more fancy maps [later](9-maps.html).

