# Maps

```{r, echo=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(fig.width = 5, fig.height = 5, fig.cap = '', collapse = TRUE)
library(terra)
#library(dismo)
#library(XML)
```

Like for other plots, there are different approaches in R to make maps. You can use "base plot" in many cases. Alternatively use levelplot, either via the `spplot` function (implemented in `sp` and `raster`) or via the `rasterVis` package.

Here are some brief examples about making maps. You can also look elsewhere on the Internet, [like here](http://www.nickeubank.com/wp-content/uploads/2015/10/RGIS3_MakingMaps_part1_mappingVectorData.html), or this for [spplot](https://edzer.github.io/sp/) and [rasterVis](https://oscarperpinan.github.io/rastervis/).


## Vector data

### Base plots 

```{r, maps_1}
library(terra)
p <- vect(system.file("exdata/lux.shp", package="terra"))
plot(p)
```

```{r, maps_2}
n <- size(p)
plot(p, col=rainbow(n))
```

One colour per region (NAME_1)

```{r, maps_3}
u <- unique(p$NAME_1)
u
m <- match(p$NAME_1, u)
plot(p, col=rainbow(n)[m])
#text(p, 'NAME_2', cex=.75, halo=TRUE)
```

### spplot

```{r, maps_4}
library(raster)
spplot(p, "AREA")
```


## Raster

Example data

```{r}
library(terra)
b <- rast(system.file("exdata/logo.grd", package="terra"))

r <- rast(p, res=0.01 )
values(r) <- 1:ncell(r)
r <- mask(r, p)
```

Several generic functions have been implemented for SpatRaster objects to create maps and other plot types. Use 'plot' to create a map of a SpatRaster. You can zoom in using 'zoom' and clicking on the map twice (to indicate where to zoom to). With `click` it is possible to interactively query a Raster* object by clicking once or several times on a map plot.

After plotting a `SpatRaster` you can add vector type spatial data (points, lines, polygons). You can do this with functions points, lines, polygons if you are using the basic R data structures or plot(object, add=TRUE) if you are using Spatial* objects as defined in the sp package. When plot is used with a multi-layer Raster* object, all layers are plotted (up to 16), unless the layers desired are indicated with an additional argument. 

```{r, maps_5}
plot(r)
plot(p, add=TRUE)
```

`image` does not provide a legend and that can be advantageous in some cases.

```{r, maps_6}
image(r)
plot(p, add=TRUE)
```

Multi-layer Raster objects can be plotted as individual layers
```{r, raster-20a, fig.width=7}
plot(b)
```

They can also be combined into a single image, by assigning individual layers to one of the three color channels (red, green and blue):
```{r, raster-20b }
plotRGB(b, r=1, g=2, b=3)
```


You can also plot `Raster*` objects with spplot. 

```{r, maps_7}
bounds <- list("sp.polygons", p)
spplot(r, sp.layout=bounds)
```


```{r, raster-20c, fig.width=9, fig.height=3}
spplot(b, layout=c(3,1))
```


The `rasterVis` package has several other `lattice` based plotting functions for `Raster*` objects. The [rasterVis](https://oscarperpinan.github.io/rastervis/) package also facilitates creating a map from a `SpatRaster` with the `ggplot2` package.


You can also use a number of other plotting functions with `raster` objects, including `hist`, `persp`, `contour`}, and `density`. See the help files for more info. 




## Specialized packages

coming soon....
