# 13. Statistical models

There are many type of statistical models. Here we show how to make simple regression models with *R*. Other modeling approaches tend to use similar syntax. 

The most common way to specify a regression model in *R* is by creating a formula. For example `y ~ x` means `y` is a function of `x`. `y ~ a + b` means that `y` is a function of `a` and `b`. 

Let's use the cars data that come with *R*. This dataset has measurements on the distance needed to stop given the speed a car was driven when the driver stepped on the breaks. We use the `lm` (linear model) function.

```{r}
head(cars)
m <- lm(dist ~ speed, data=cars)
m
```

Note that the data is provided by data.frame `cars`, and that the names in formula are column names in this `data.frame`. The functions returned a model (`lm`) object. When printed it shows the coefficients of the regression model (`dist = -17.579 + 3.932 * speed`). `m` has quite a bit more information, but that is not shown, by default.

There are several functions that can be used to extract this information. 

```{r}
summary(m)
anova(m)
residuals(m)[1:10]
```

You can use `abline` to draw a simple regression line like this.

```{r, stat1, fig.cap=''}
plot(cars, col='blue', pch='*', cex=2)
abline(m, col='red', lwd=2)
```

More generally, you can use the `predict` function to use the model to predict values of `y` for any `x`.

```{r, stat2, fig.cap=''}
p <- predict(m, data.frame(speed=1:30))
p
plot(1:30, p, xlab='speed', ylab='distance', type='l', lwd=2)
points(cars)
```

The `glm` (generalized linear models) function can do what `lm` can, but it is much more versatile. For example you can also use it for logistic regression. In logistic regression the response variable is normally binomial (0 or 1) or at least between 0 and 1. I create such a variable here (was the stopping distance above 40 or not?). 

```{r}
cars$above40 <- cars$dist > 40
```

Now we can use this variable in a `glm` model. By stating that `family='binomial'` we indicate that we want logistic regression. (The default is `family=gaussian` which indicates standard (normal) regression.

```{r}
mlog <- glm(above40 ~ speed, data=cars, family='binomial')
mlog
```

Because a logistic model operates with logistically transformed numbers, we need to tell the predict function that we want the predicted values on the original scale (`type='response'`). 

```{r}
p <- predict(mlog, data.frame(speed=1:30), type='response')
```

```{r, stat10, fig.cap=''}
plot(cars$speed, cars$above40)
lines(1:30, p)
```

For a more in-depth discussion of regression models with R [go here](http://tutorials.iq.harvard.edu/R/Rstatistics/Rstatistics.html).
