# Algebra

Vectors and matrices can be used to compute new vectors (matrices) with simple and intuitive algebraic expressions. 

## Vector algebra

We have two vectors, `a` and `b`

```{r}
a <- 1:5
b <- 6:10
```

Multiplication works element by element. That is `a[1] * b[1]`, `a[2] * b[2]`, etc
```{r}
d <- a * b
a
b
d
```

The examples above illustrate a special feature of *R* not found in most other programming languages. This is that you do not need to 'loop' over elements in an array (vector in this case) to compute new values. It is important to use this feature as much as possible. In other programming languages you would need to do something like the 'for-loop' below to achieve the above (for-loops do exist in *R* and are discussed in a later chapter). 

You can also multiply with a single number. 
```{r}
a * 3
```

In the examples above the computations used either vectors of the same length, or one of the vectors had length 1. But be careful, you can use algebraic computations with vectors of different lengths, as the shorter ones will be "recycled". *R* only issues a warning if the length of the longer vector is not a multiple of the length of the shorter object. This is a great feature when you need it, but it may also make you overlook errors when your data are not what you think they are.

```{r}
a + c(1,10)
```

No warning here:
```{r}
1:6 + c(0,10)
```


### Logical comparisons 

Recall that `==` is used to test for equality
```{r}
a == 2
f <- a > 2
f
```

`&` is Boolean "AND", and `|` is Boolean "OR". 

```{r}
a
b
b > 6 & b < 8
# combining a and b
b > 9 | a < 2
```

"Less than or equal" is `<=`, and "more than or equal" is `>=`.  
```{r}
b >= 9
a <= 2
b >= 9 | a <= 2
b >= 9 & a <= 2
```


### Functions

There are many functions that allow us to do vectorized algebra. For example:

```{r}
sqrt(a)
exp(a)
```

Not all functions return a vector of the same length. The following functions return just one or two numbers:
```{r}
min(a)
max(a)
range(a)
sum(a)
mean(a)
median(a)
prod(a)
sd(a)
```

If you cannot guess what `prod` and `sd` do, look it up in the help files (e.g. `?sd`)


### Random numbers

It is common to create a vector of random numbers in data analysis, and also to create example data to demonstrate how a procedure works. To get 10 numbers sampled from the uniform distribution between 0 and 1 you can do
```{r}
r <- runif(10)
r
```

For Normally distributed numbers, use `rnorm`

```{r}
r <- rnorm(10, mean=10, sd=2)
r
```

If you run the functions above, you will get different numbers. After all, they are random numbers! Well, computer generated numbers are not truly random, but 'pseudo-random'. To be able to exactly reproduce examples or data analysis we often want to assure that we take exactly the same "random" sample each time we run our code. To do that we use `set.seed`. This function initializes the random number generator to a specific point. This is illustrated below.

```{r}
set.seed(12)
runif(3)
runif(4)
runif(5)

set.seed(12)
runif(3)
runif(5)

set.seed(12)
runif(3)
runif(5)
```

Note that each time `set.seed` is called, the same sequence of (pseudo) random numbers will be generated. This is a very important feature, as it allows us to exactly reproduce results that involve random sampling. The seed number is arbitrary; a different seed number will give a different sequence.

```{r}
set.seed(12)
runif(3)
runif(5)
```

## Matrices

Computation with matrices is also 'vectorized'. For example, with matrix `m` you can do `m * 5` to multiply all values of m3 with 5, or do `m^2` or `m * m` to square the values of m. 

```{r}
# set up an example matrix
m <- matrix(1:6, ncol=3, nrow=2, byrow=TRUE)
m

m * 2

m^2
```

We can also do math with a matrix and a vector. Note, again, that computation with matrices in *R* is column-wise, and that shorter vectors are recycled. 

```{r}
m * 1:2
```

You can multiply two matrices.
```{r}
m * m
```
Note that this is "cell by cell" multiplication. For ['matrix multiplication'](https://en.wikipedia.org/wiki/Matrix_multiplication) in the mathematical sense, you need to use the `%*%` operator.

```{r}
m %*% t(m)
```
