# Data exploration

After reading data from a file (see the previous chapter), the next thing to do is to look at some summary statistics. This is in the first place just to check your data. Very often you discover some strange values that are not quite right. Careful inspection of your data after you read it is very important. It can help avoid a lot of trouble later on when you are trying to explain odd results!

Sometimes you can correct errors in *R* but in other cases you first need to fix the data file. Fixing a data file can be the way to go if you are dealing with your own primary data. However when you are working with a file provided by someone else it is generally better to not change the file, but to correct things via *R* code, if possible. That leaves an exact trail of what you have done, and allows you to apply the same corrections to a new version of the data. 

Inspecting your data is also important to understand it better, and such "exploratory data analysis" can be an important step before doing the specific analyses of interest. There are many general and specialized tools for this, and here we only cover some of the very basic tools. 


## Summary and table

Consider `data.frame` `d`
```{r}
d <- data.frame(id=1:10, 
        name=c('Bob', 'Bobby', '???', 'Bob', 'Bab', 'Jim', 'Jim', 'jim', '', 'Jim'), 
		score1=c(8, 10, 7, 9, 2, 5, 1, 6, 3, 4), 
		score2=c(3,4,5,-999,5,5,-999,2,3,4), stringsAsFactors=FALSE)
d
```

`d` is very small and we can simply look at the values in `d` to see if they look all-right. But with real data you may have 100s or 1000s of rows and many columns, making it very hard, if not impossible, to spot errors by eye-balling.
  
The summary function is an easy way to start, at least for numeric variables.  
```{r}
summary(d)
```

The minimum value of variable `score2` is -999. That was probably used in data entry to indicate a missing value. These should be changed to `NA`. 

```{r}
# which values in score2 are -999?
i <- d$score2 == -999
# set these to NA
d$score2[i] <- NA
summary(d)
```

The two steps used above: `i <- d$score2 == -999` and `d$score2[i] <- NA` are usually done in a single line: `d$score2[d$score2 == -999] <- NA`.

For character (and integer) variables it can be useful to use `unique` and `table`:

```{r}
unique(d$name)
table(d$name)
```

Often you will discover slight variations in spelling that need to be corrected. In this case, let's assume that 'Bobby' and 'Bab' should both be 'Bob'. We replace 'Bab' and 'Bobby' with 'Bob'.  

```{r}
d$name[d$name %in% c('Bab', 'Bobby')] <- 'Bob'
table(d$name)
```

'jim' should be "Jim". It is easy enough to replace as done above. But what if there were many cases like that? 
It would be easy to make all character values lower- or uppercase with `d$name <- toupper(d$name)` but I want to keep the normal name capitalization of the first letter only. So let's assure that all names start with an uppercase letter.

```{r}
# get the first letters
first <- substr(d$name, 1, 1)
# get the remainder
remainder <- substr(d$name, 2, nchar(d$name))
# assure that the first letter is upper case
first <- toupper(first)
# combine
name <- paste0(first, remainder)
# assign back to the variable
d$name <- name
table(d$name)
```

The question marks in `name` should probably also be replaced with `NA`.

```{r}
d$name[d$name == '???'] <- NA
table(d$name)
```

You can force `table` to also count the `NA` values:
```{r}
table(d$name, useNA='ifany')
```

Note that there is one 'empty' value.

```{r}
d$name[9]
```

That should also be a missing value in this case.

```{r}
d$name[d$name == ''] <- NA
table(d$name, useNA='ifany')
```

You can also use `table` to make a contingency table of two variables.
```{r}
table(d[ c('name', 'score2')])
``` 

## Quantile, range, and mean

Other useful functions include `quantile`, `range`, and `mean`.

```{r}
quantile(d$score1)
range(d$score1)
mean(d$score1)
```

Note that in some functions you may need to use `na.rm=TRUE` if there are `NA` values. Otherwise, as soon as there is a single `NA` value in a computation, the results becomes `NA`. This is very common in *R* ---  so keep that in mind if all your results are `NA`. 

```{r}
quantile(d$score2)
range(d$score2)
quantile(d$score2, na.rm=TRUE)
range(d$score2, na.rm=TRUE)
```

In this data exploration phase it is very useful to make plots. We'll discuss plotting in a later chapter, but here are four example plots. Note how `par(mfrow=c(2,2))` sets up the canvas for two rows and columns, that is for four plots.
 
```{r, plot1, fig.cap=""}
par(mfrow=c(2,2))
plot(d$score1, d$score2)
boxplot(d[, c('score1', 'score2')])
plot(sort(d$score1))
hist(d$score2)
```
