# Apply

The "apply family" of functions (`apply`, `tapply`, `lapply` and others) and related functions such as `aggregate` are central to using *R*. They provide an concise, elegant and efficient approach to apply (sometimes referred to as "to map") a function to a set of cases, be they rows or columns in a matrix or data.frame, or elements in a list. 

## apply

Consider matrix `m`

```{r}
m <- matrix(1:15, nrow=5, ncol=3)
m
```

## apply

Computation with matrices is 'vectorized'. For example you can do `m * 5`to multiply all values of m with 5 or do `m^2` or `m * m` to square the values of `m`. But often we need to compute values for the margins of a matrix, that is, a single value for each row or column. The `apply` function can be used for that:

```{r}
# sum values in each row
apply(m, 1, sum)

# get mean for each column
apply(m, 2, mean)
```

Note that the apply uses at least three arguments: a matrix, a 1 or 2 indicating whether the computation is for rows or for columns, and a function that computes a new value (or values) for each row or column. You can read more about this in the help file of the function (type `?apply` ). In most cases you will also add the argument `na.rm=TRUE` to remove `NA` (missing) values as any computation that includes an `NA` value will return `NA`. In this case we used existing basic functions `mean` and `sum` but we could also supply a function that we wrote ourselves. 

Note that `apply`(and related functions such as `tapply` and `sapply` are ways to avoid writing a loop. In the `apply` examples above you could have written a loop to do the computations row by row (or column by column) but using `apply` is more compact and efficient.

The rowSums and colSums functions are (fast) shorthand functions for apply( , , sum)

```{r}
rowSums(m)
```


## tapply


`tapply` can be used to compute a summary statistic, e.g. a mean value, for groups of rows in a data.frame. You need one column that indicates the group, and then you can compute, for example, the mean value for that group.

```{r}
colnames(m) <- c('v1', 'v2', 'v3')
d <- data.frame(name=c('Yi', 'Yi', 'Yi', 'Er', 'Er'), m, stringsAsFactors=FALSE)
d$v2[1] <- NA
d
``` 

Imagine that you would like to compute the average value of `v1`, `v2` and `v3` for each individual (`name`). You can use `tapply` for that.

```{r}
tapply(d$v1, d$name, mean)
tapply(d$v1, d$name, max)
tapply(d$v2, d$name, mean)
tapply(d$v2, d$name, mean, na.rm=TRUE)
```

## aggregate

`aggregate` is similar to `tapply` but more convenient if you want to compute a summary statistic for multiple variables. It does have the annoying problem that the second argument cannot be a vector:

```{r}
aggregate(d[, c('v1', 'v2', 'v3')], d$name, mean, na.rm=TRUE)
```

You can fix that in two ways

```{r}
aggregate(d[, c('v1', 'v2', 'v3')], d[, 'name', drop=FALSE], mean, na.rm=TRUE)
# or 
aggregate(d[, c('v1', 'v2', 'v3')], list(d$name), mean, na.rm=TRUE)
```

As explained before, this is why the first one works: when you extract a single column from a `matrix` or `data.frame`, the structure (class) "drops" to a simpler form, it becomes a vector. `drop=FALSE` stops that from happening. 


## sapply and lapply

To iterate over a list, we can use `lapply` or `sapply` . The difference is that `lapply` always returns a list while `sapply` tries to simplify the result to a vector or matrix.

```{r}
names <- list('Antoinette', 'Mary', 'Duncan', 'Obalaya', 'Jojo')
nchar('Jim')

lapply(names, nchar)
sapply(names, nchar)
```

In all cases `(t)(s)(l)apply` and `aggregate` (and many more functions) we provided some data and a function, such as `mean` or `nchar`.
You can also provide your own custom function. For example

```{r}
shortname <- function(name) {
	if (nchar(name) < 5) {
		name <- toupper(name)
		return(name)
	} else {
		name <- substr(name,1,5)
		return(name)
	}
}

sapply(names, shortname)
```

More examples: https://nsaunders.wordpress.com/2010/08/20/a-brief-introduction-to-apply-in-r/


